---
title: "DataAnalysis"
author: "Nora Abbott"
date: "2023-04-05"
output: pdf_document
---

```{r library}
# load packages
library(ggplot2)
library(ggthemes)
library(RColorBrewer)
library(moments) 
library(rstatix)
library(kableExtra)
library(dplyr)
library(ggpubr)
library(tidyverse)
library(coin)
library(gridExtra)
library(lmtest)
library(GGally)
library(ggformula)
library(ggpmisc)
library(car)
library(readxl)
library(vtable)
library(gridExtra)
library(cowplot)
library(gvlma)
library(report)
library(ggstance)
library(jtools)
library(interactions)
library(psych)
library(kableExtra)
library(gtools)
library(caret)
library(performance)
library(MASS)
library(lmtest)
library(sandwich)
library(ggplot2)  
library(lattice) 
library(blorr) 
library(glmtoolbox)
library(lmtest)  
library(pROC)
library(arm)
library(pROC)
library(repmod)
library(countreg)
library(tidymodels)
library(tidyflow)
library(rpart.plot)
library(vip)
library(baguette)
library(ranger)
library(xgboost)
library(sf)
library(tidycensus)
library(viridis)
library(tigris)
library(usmap)
library(rgdal)
```

# DATA CLEANING AND PREPARATION

```{r}
# read census block group data from GitHub
blk_data <- read.csv("https://raw.githubusercontent.com/ricerb/biol-562-project/master/data/biol562%20project%20dataset%20v1.csv")
```

```{r}
# select required blk group variables 
select_data <- blk_data[,c("GEOID", "STATEFP", "ADI_NATRANK", "STATE","POPULATION","TRACTA")]

# set empty cells to NA and omit
select_data[select_data == ""] <- NA   

# rename total population variable
names(select_data)[names(select_data) == "POPULATION"] <- "BLK_POP.ppl"
select_data$BLK_POP.ppl <- as.numeric(as.character(select_data$BLK_POP.ppl))
```

```{r}
# rename block GEOID variable
names(select_data)[names(select_data) == "GEOID"] <- "GEOID_blk"

# add back leading zeros onto census data
select_data$GEOID_blk <- as.character(select_data$GEOID_blk)
select_data$GEOID_blk <- str_pad(select_data$GEOID_blk, pad="0", width = 12, side="left")
# create new variable for corresponding tract GEOID
select_data$GEOID_tract <- substr(select_data$GEOID_blk,1,11)
```

```{r}
# calculate block group population-weighted NATL API index scores for census tracts
select_data$ADI_NATRANK <- as.numeric(as.character(select_data$ADI_NATRANK))
select_data <- na.omit(select_data)

select_data <- select_data %>% group_by(GEOID_tract) %>% mutate(TRACT_POP.ppl = sum(BLK_POP.ppl))
select_data$BLK_wt <- select_data$BLK_POP.ppl / select_data$TRACT_POP.ppl
select_data$ADI_cont <- (select_data$BLK_wt)*(select_data$ADI_NATRANK)

select_data <- select_data %>% group_by(GEOID_tract) %>% mutate(TRACT_ADI.ave = sum(ADI_cont))
```

```{r}
# read census tract data from GitHub
tract_data <- read.csv("https://raw.githubusercontent.com/ricerb/biol-562-project/master/data/biol562%20project%20dataset%20census%20tract%20level%20v1.csv")
```

```{r}
# select census tract variables of interest
data <- tract_data[,c("GEOID", "TRACTA","STATEFP","STATE", "STUSAB","COUNTY","YEAR", "pm25", "AREALAND", "AREAWATR","H7V001","H7X002","H7X003","H7X004","H7X005","H7X006","JN9E015","JN9E032","H7W002","H7W005", "JQBE001")]

# replace empty cells with NA and remove
data[data == ""] <- NA   
data <- na.omit(data)

# rename variables in tract data frame
names(data)[names(data) == "pm25"] <- "PM2.5.ug_m3_LC"
names(data)[names(data) == "AREALAND"] <- "AREALAND.m2"
names(data)[names(data) == "AREAWATR"] <- "AREAWATR.m2"
names(data)[names(data) == "H7V001"] <- "TOTAL_POP.ppl"
names(data)[names(data) == "H7W002"] <- "URBAN_POP.ppl"
names(data)[names(data) == "H7W005"] <- "RURAL_POP.ppl"
names(data)[names(data) == "STUSAB"] <- "STATE_abbr"
names(data)[names(data) == "H7X002"] <- "POPWH.ppl"
names(data)[names(data) == "H7X003"] <- "POPBLKAA.ppl"
names(data)[names(data) == "H7X004"] <- "POPNATIVE.ppl"
names(data)[names(data) == "H7X005"] <- "POPASIAN.ppl"
names(data)[names(data) == "H7X006"] <- "POPNATIVEISL.ppl"
names(data)[names(data) == "JN9E015"] <- "BACHDEG_MALE25.ppl"
names(data)[names(data) == "JN9E032"] <- "BACHDEG_FEM25.ppl"
names(data)[names(data) == "JQBE001"] <- "PERCAPINCOME_12mo.2010USD"
```

```{r}
# calculate population density. (From Sage-) I LOOKED THIS UP, pop density is people/sq units of land. also our stuff looked weird with this as sq meter so i made it sq kilo.
blkGrpA <- data$AREALAND.m2
blkGrpPopDen <- data$TOTAL_POP.ppl / blkGrpA
data$POPDENS.ppl_km2 <- blkGrpPopDen *10^6

# calculate census tract proportions
data$PROP_URBAN.percent <- (data$URBAN_POP.ppl / data$TOTAL_POP.ppl) *100
data$PROP_RURAL.percent <- (data$RURAL_POP.ppl / data$TOTAL_POP.ppl) *100
data$PROP_WH.percent <- (data$POPWH.ppl / data$TOTAL_POP.ppl) *100
data$PROP_BLKAA.percent <- (data$POPBLKAA.ppl / data$TOTAL_POP.ppl) *100
data$PROP_NATIVE.percent <- (data$POPNATIVE.ppl / data$TOTAL_POP.ppl) *100
data$PROP_ASIAN.percent <- (data$POPASIAN.ppl / data$TOTAL_POP.ppl) *100
data$PROP_NATIVEISL.percent <- (data$POPNATIVEISL.ppl / data$TOTAL_POP.ppl) *100
data$PROP_BACHDEG_25.percent <- ((data$BACHDEG_MALE25.ppl + data$BACHDEG_FEM25.ppl) / data$TOTAL_POP.ppl) *100
data$log_POPDENS.ln_ppl_km2 <- log(data$POPDENS.ppl_km2)
```

```{r}
# replace % degree holders greater than 100 with NAN
is.na(data$PROP_BACHDEG_25.percent) <- data$PROP_BACHDEG_25.percent > 100
```

```{r}
# add back leading zeros onto census data
data$GEOID <- as.character(data$GEOID)
data$GEOID <- str_pad(data$GEOID, pad="0", width = 11, side="left")
```

```{r}
# merge population weighted census blk ADIs with census tract data
blk_data_select <- select_data[,c("GEOID_tract","TRACT_ADI.ave")]
blk_data_select <- distinct(blk_data_select)
names(blk_data_select)[names(blk_data_select) == "GEOID_tract"] <- "GEOID"

data <- left_join(data,blk_data_select,by="GEOID")
```

```{r}
# determine PM2.5 quantiles and assign quantile numbers to census tracts
pm25_quant <- quantile(data$PM2.5.ug_m3_LC, na.rm=TRUE)
pm25_quant

pm25_q1 <- pm25_quant[2]
pm25_q2 <- pm25_quant[3]
pm25_q3 <- pm25_quant[4]
pm25_q4 <- pm25_quant[5]

data_pm25_Q1 <- data[data$PM2.5.ug_m3_LC < pm25_q1,]
data_pm25_Q2 <- data[data$PM2.5.ug_m3_LC >= pm25_q1 & data$PM2.5.ug_m3_LC < pm25_q2,]
data_pm25_Q3 <- data[data$PM2.5.ug_m3_LC >= pm25_q2 & data$PM2.5.ug_m3_LC < pm25_q3,]
data_pm25_Q4 <- data[data$PM2.5.ug_m3_LC >= pm25_q3,]

data$PM25_quant <- as.factor(ifelse(data$PM2.5.ug_m3_LC<pm25_q1, '1',
                     ifelse(data$PM2.5.ug_m3_LC<pm25_q2, '2',
                     ifelse(data$PM2.5.ug_m3_LC<pm25_q3, '3',
                     ifelse(data$PM2.5.ug_m3_LC<=pm25_q4, '4', 'E')))))
```

```{r}
# CENSUS DEFINE REGIONS
NE_reg <- c("CT","NJ","NY","PA","MA","NH","VT","RI","ME")

MW_reg <- c("IL","IN","MI","OH","WI","IA","KS","MN","MO","NE","ND","SD")

S_reg <- c("DE","DC","FL","GA","MD","NC","SC","VA","WV","AL","KY","MS","TN","AR","LA","OK","TX")

W_reg <- c("AZ","CO","ID","MT","NV","NM","UT","WY","CA","OR","WA")
```

```{r}
# assign tracts regions
data$CENSUS_REG <- as.factor(ifelse(data$STATE_abbr %in% NE_reg, 'Northeast',
                     ifelse(data$STATE_abbr %in% MW_reg, 'Midwest',
                     ifelse(data$STATE_abbr %in% S_reg, 'South',
                     ifelse(data$STATE_abbr %in% W_reg, 'West', 'Other')))))
```

# SUMMARY STATISTICS
```{r}
data %>% summarize(mean=mean(PM2.5.ug_m3_LC), sd=sd(PM2.5.ug_m3_LC), median=median(PM2.5.ug_m3_LC), n=n())

sum(data$TOTAL_POP.ppl)

# primary nonattainment
sum(data$PM2.5.ug_m3_LC > 12)
sum(data$PM2.5.ug_m3_LC > 12)/71857

# secondary nontaainment
sum(data$PM2.5.ug_m3_LC > 15)
sum(data$PM2.5.ug_m3_LC > 15)/71857
```

# EXPLORATORY PLOTS

```{r}
#ggpairs(data, columns= c("PM2.5.ug_m3_LC","TRACT_ADI.ave","POPDENS.ppl_km2","PROP_URBAN.percent","PROP_BACHDEG_25.percent","PROP_BLKAA.percent","PERCAPINCOME_12mo.2010USD"), alpha=0.2)
```

```{r}
# histograms for PM2.5
pmHist<- ggplot(data, aes(x=PM2.5.ug_m3_LC)) + geom_histogram(bins=60, color="lightgrey", fill="lightgrey") + theme_classic() + ylab("Count") + xlab("Average PM2.5 in Census Tracts (ug/m3 (LC))") +geom_vline(xintercept = 12, color="darkred", size=1)+ 
  geom_text(aes(x=15, label="EPA Secondary Annual Average Standard = 15 ug/m3 \n", y=2200), color="darkorange", angle=90, text=element_text(size=11))+
  geom_vline(xintercept = 15, color="darkorange", size=1)+ 
  geom_text(aes(x=12, label="EPA Primary Annual Average Standard = 12 ug/m3 \n", y=2200), color="darkred", angle=90, text=element_text(size=11))
pmHist
ggsave("pmHist.png",pmHist)
```

```{r}
# boxplots from pm2.5 quantiles
pmBxp<-ggplot(data=data,aes(x=PM25_quant,y=PM2.5.ug_m3_LC,group=PM25_quant)) + geom_boxplot(aes(fill=PM25_quant))+ theme_classic() + xlab("PM2.5 Quantile") + ylab("PM2.5 Concentration (ug/m3)") + scale_fill_brewer(palette="Dark2", name="PM2.5 Quantile")
ggsave("pmBxp.png",pmBxp)
```

```{r}
# boxplots from pm2.5 quantiles vs national AQI
adiBxp<-ggplot(data=data,aes(x=PM25_quant,y=TRACT_ADI.ave,group=PM25_quant)) + geom_boxplot(aes(fill=PM25_quant))+ theme_classic() + xlab("PM2.5 Quantile") + ylab("ADI National Index Score")+ scale_fill_brewer(palette="Dark2", name="PM2.5 Quantile") + coord_flip()
ggsave("adiBxp.png",adiBxp)
```

```{r, fig.width=7,fig.height=7}
# scatterplot from pm2.5 vs national ADI
adiScat<-ggplot(data=data,aes(y=PM2.5.ug_m3_LC,x=TRACT_ADI.ave)) + geom_point(alpha=0.05) + theme_classic() + ylab("PM2.5 (ug/m3)") + xlab("ADI National Index Score") + geom_smooth(method=lm, color="blue") + theme(axis.title.x = element_text(size = 14),axis.title.y = element_text(size = 14))
adiScat
ggsave("adiScat.png", adiScat)
```

```{r}
# boxplots from pm2.5 quantiles vs population density
popdensBxp<-ggplot(data=data,aes(x=PM25_quant,y=POPDENS.ppl_km2,group=PM25_quant)) + geom_boxplot(aes(fill=PM25_quant))+ theme_classic() + xlab("PM2.5 Quantile") + ylab("Population Density (people/km2)") + scale_fill_brewer(palette="Dark2", name="PM2.5 Quantile") + coord_flip() 
ggsave("popdensBxp.png",popdensBxp)
```

```{r, fig.width=7,fig.height=7}
# scatterplot from popDens vs pm2.5
popdensScat<-ggplot(data=data,aes(x=POPDENS.ppl_km2,y=PM2.5.ug_m3_LC)) + geom_point(alpha=0.05) + theme_classic() + ylab("PM2.5 (ug/m3)") + xlab("Population Density (people/km2)") + geom_smooth(method=lm, color="blue") + ylim(0, 25) + theme(axis.title.x = element_text(size = 14),axis.title.y = element_text(size = 14))
ggsave("popdensScat.png",popdensScat)
```

```{r}
# boxplots from pm2.5 quantiles vs log population density
popdensBxp<-ggplot(data=data,aes(x=PM25_quant,y=log_POPDENS.ln_ppl_km2,group=PM25_quant)) + geom_boxplot(aes(fill=PM25_quant))+ theme_classic() + xlab("PM2.5 Quantile") + ylab("Ln(Population Density) (ln(people/km2))") + scale_fill_brewer(palette="Dark2", name="PM2.5 Quantile") + coord_flip() 
ggsave("logpopdensBxp.png",popdensBxp)
```

```{r, fig.width=7,fig.height=7}
# scatterplot from log popDens vs pm2.5
popdensScat<-ggplot(data=data,aes(x=log_POPDENS.ln_ppl_km2,y=PM2.5.ug_m3_LC)) + geom_point(alpha=0.05) + theme_classic() + ylab("PM2.5 (ug/m3)") + xlab("Ln(Population Density) (ln(people/km2))") + geom_smooth(method=lm, color="blue") + ylim(0, 25) + theme(axis.title.x = element_text(size = 14),axis.title.y = element_text(size = 14))
ggsave("logpopdensScat.png",popdensScat)
```

```{r}
# boxplots from pm2.5 quantiles vs % urban
urbanBxp<-ggplot(data=data,aes(x=PM25_quant,y=PROP_URBAN.percent,group=PM25_quant)) + geom_boxplot(aes(fill=PM25_quant))+ theme_classic() + xlab("PM2.5 Quantile") + ylab("Percent Urban Population (%)") + scale_fill_brewer(palette="Dark2", name="PM2.5 Quantile") + coord_flip()
ggsave("urbanBxp.png",urbanBxp)
```

```{r, fig.width=7,fig.height=7}
# scatterplot from % urban vs pm2.5
urbanScat<-ggplot(data=data,aes(x=PROP_URBAN.percent,y=PM2.5.ug_m3_LC)) + geom_point(alpha=0.05) + theme_classic() + ylab("PM2.5 (ug/m3)") + xlab("Percent Urban Population (%)") + geom_smooth(method=lm, color="blue") + theme(axis.title.x = element_text(size = 14),axis.title.y = element_text(size = 14))
ggsave("urbanScat.png",urbanScat)
```

```{r}
# boxplots from pm2.5 quantiles vs % rural
ruralBxp<-ggplot(data=data,aes(x=PM25_quant,y=PROP_RURAL.percent,group=PM25_quant)) + geom_boxplot(aes(fill=PM25_quant))+ theme_classic() + xlab("PM2.5 Quantile") + ylab("Percent Rural Population (%)") + scale_fill_brewer(palette="Dark2", name="PM2.5 Quantile") + coord_flip()
ggsave("ruralBxp.png",ruralBxp)
```

```{r}
# boxplots from pm2.5 quantiles vs % 25+ with bachelor degree
bachdegBxp<-ggplot(data=data,aes(x=PM25_quant,y=PROP_BACHDEG_25.percent,group=PM25_quant)) + geom_boxplot(aes(fill=PM25_quant))+ theme_classic() + xlab("PM2.5 Quantile") + ylab("Percent Population 25+ with Bachelors (%)") + scale_fill_brewer(palette="Dark2", name="PM2.5 Quantile") + coord_flip()
ggsave("bachdegBxp.png",bachdegBxp)
```

```{r, fig.width=7,fig.height=7}
# scatterplot from % bachelors vs pm2.5
bachdegScat<-ggplot(data=data,aes(x=PROP_BACHDEG_25.percent,y=PM2.5.ug_m3_LC)) + geom_point(alpha=0.05) + theme_classic() + ylab("PM2.5 (ug/m3)") + xlab("Percent Population 25+ with Bachelors (%)")+ geom_smooth(method=lm, color="blue") + theme(axis.title.x = element_text(size = 14),axis.title.y = element_text(size = 14))
ggsave("bachdegScat.png",bachdegScat)
```

```{r}
# boxplots from pm2.5 quantiles vs % Black/African American
BAAbxp<-ggplot(data=data,aes(x=PM25_quant,y=PROP_BLKAA.percent,group=PM25_quant)) + geom_boxplot(aes(fill=PM25_quant))+ theme_classic() + xlab("PM2.5 Quantile") + ylab("Percent Black or African American (%)")+ scale_fill_brewer(palette="Dark2", name="PM2.5 Quantile") + coord_flip()
ggsave("BAAbxp.png",BAAbxp)
```

```{r, fig.width=7,fig.height=7}
# scatterplot from % Black or African American vs pm2.5
BAAscat<-ggplot(data=data,aes(x=PROP_BLKAA.percent,y=PM2.5.ug_m3_LC)) + geom_point(alpha=0.05) + theme_classic() + ylab("PM2.5 (ug/m3)") + xlab("Percent Black or African American Only (%)")+ geom_smooth(method=lm, color="blue") + theme(axis.title.x = element_text(size = 14),axis.title.y = element_text(size = 14))
ggsave("BAAscat.png",BAAscat)
```

```{r, fig.width=7,fig.height=7}
# scatterplot from % White only vs pm2.5
whScat<-ggplot(data=data,aes(x=PROP_WH.percent,y=PM2.5.ug_m3_LC)) + geom_point(alpha=0.05) + theme_classic() + ylab("PM2.5 (ug/m3)") + xlab("Percent White Only (%)")+ geom_smooth(method=lm, color="blue") + theme(axis.title.x = element_text(size = 14),axis.title.y = element_text(size = 14))
ggsave("whScat.png",whScat)
```

```{r, fig.width=7,fig.height=7}
# scatterplot from % Asian vs pm2.5
AsScat<-ggplot(data=data,aes(x=PROP_ASIAN.percent,y=PM2.5.ug_m3_LC)) + geom_point(alpha=0.05) + theme_classic() + ylab("PM2.5 (ug/m3)") + xlab("Percent Asian Only (%)")+ geom_smooth(method=lm, color="blue") + theme(axis.title.x = element_text(size = 14),axis.title.y = element_text(size = 14))
ggsave("AsScat.png",AsScat)
```

```{r, fig.width=7,fig.height=7}
# scatterplot from % Native vs pm2.5
NAscat<-ggplot(data=data,aes(x=PROP_NATIVE.percent,y=PM2.5.ug_m3_LC)) + geom_point(alpha=0.05) + theme_classic() + ylab("PM2.5 (ug/m3)") + xlab("Percent American Indian and Alaska Native Alone")+ geom_smooth(method=lm, color="blue") + theme(axis.title.x = element_text(size = 14),axis.title.y = element_text(size = 14))
ggsave("NAscat.png",NAscat)
```

```{r, fig.width=7,fig.height=7}
# scatterplot from % Native Hawiann or Pacific Islander vs pm2.5
NHPIscat<-ggplot(data=data,aes(x=PROP_NATIVEISL.percent,y=PM2.5.ug_m3_LC)) + geom_point(alpha=0.05) + theme_classic() + ylab("PM2.5 (ug/m3)") + xlab("Percent Native Hawaiian and Other Pacific Islander Alone")+ geom_smooth(method=lm, color="blue") +ylim(0, NA) + theme(axis.title.x = element_text(size = 14),axis.title.y = element_text(size = 14))
ggsave("NHPIscat.png",NHPIscat)
```

```{r}
# boxplots from pm2.5 quantiles vs Per CAP INCOME
incBxp<-ggplot(data=data,aes(x=PM25_quant,y=PERCAPINCOME_12mo.2010USD,group=PM25_quant)) + geom_boxplot(aes(fill=PM25_quant))+ theme_classic() + xlab("PM2.5 Quantile") + ylab("12mo Per Capita Income (2010 USD)")+ scale_fill_brewer(palette="Dark2", name="PM2.5 Quantile") + scale_y_continuous(labels = scales::comma) + coord_flip()
ggsave("incBxp.png",incBxp)
```

```{r, fig.width=7,fig.height=7}
# scatterplot for Per CAP INCOME vs pm2.5
incScat<-ggplot(data=data,aes(x=PERCAPINCOME_12mo.2010USD,y=PM2.5.ug_m3_LC)) + geom_point(alpha=0.05) + theme_classic() + ylab("PM2.5 (ug/m3)") + xlab("12mo Per Capita Income (2010 USD)") + scale_x_continuous(labels = scales::comma)+ geom_smooth(method=lm, color="blue") + theme(axis.title.x = element_text(size = 14),axis.title.y = element_text(size = 14))
ggsave("incScat.png",incScat)
```

# EXPLORATORY MAPS

```{r}
test <- read_sf("tractShpfile/US_tract_2010.shp")

UStract <- subset(test, !(STATEFP10 %in% c("02", "15", "72")))
```

```{r}
# merge spatial, census, and ADI data
names(UStract)[names(UStract) == "GEOID10"] <- "GEOID"

UStractMerge <- UStract[c("GEOID","geometry")]
```

```{r}
data_sf <- merge(UStractMerge,data,by="GEOID")
```

```{r}
# download states from tigris
states<- states(cb=TRUE)

# assign tracts regions to states
states$CENSUS_REG <- as.factor(ifelse(states$STUSPS %in% NE_reg, 'Northeast',
                     ifelse(states$STUSPS %in% MW_reg, 'Midwest',
                    ifelse(states$STUSPS %in% S_reg, 'South',
                     ifelse(states$STUSPS %in% W_reg, 'West', 'Other')))))

# remove other census regions
states<-states[states$CENSUS_REG != "Other",]

regionPlotState <- ggplot(states) +geom_sf(aes(fill=CENSUS_REG), color=NA)+ 
  labs(title = "Census Regions for Continental Sates")+
  theme(axis.text =  element_blank(),
    axis.ticks =  element_blank(),
    panel.background = element_blank()) + labs(fill='Census Region') 
regionPlotState
ggsave("regionPlotState.png",regionPlotState)

regionPlotTract <- ggplot(data_sf) +geom_sf(aes(fill=CENSUS_REG), color=NA)+ 
  labs(title = "Census Regions for Census Tracts included in Analysis")+
  theme(axis.text =  element_blank(),
    axis.ticks =  element_blank(),
    panel.background = element_blank()) + labs(fill='Census Region') 
regionPlotTract
ggsave("regionPlotTract.png",regionPlotTract)
```

```{r}
pmPlot<-ggplot(data_sf)+geom_sf(aes(fill=PM2.5.ug_m3_LC), color = NA) + 
  labs(title = "Average PM2.5 in US Census Tracts") + 
  scale_fill_gradient(low = "grey94", high = "sienna2", na.value ="darkgrey", name="Concentration (ug/m3)")+
  theme(axis.text =  element_blank(),
    axis.ticks =  element_blank(),
    panel.background = element_blank())
ggsave("pmPlot.png",pmPlot)
```

```{r}
whitePlot<-ggplot(data_sf)+geom_sf(aes(fill=PROP_WH.percent), color = NA) + 
  labs(title = "Percent White Alone in US Census Tracts") + 
  scale_fill_gradient(low = "grey94", high = "tomato", na.value ="darkgrey", name="Percent (%)") +  
  theme(axis.text =  element_blank(),
    axis.ticks =  element_blank(),
    panel.background = element_blank())
ggsave("whitePlot.png",whitePlot)
```

```{r}
blkAaPlot<-ggplot(data_sf)+geom_sf(aes(fill=PROP_BLKAA.percent), color = NA) + 
  labs(title = "Percent Black or African American Alone in US Census Tracts") + 
  scale_fill_gradient(low = "grey94", high = "slateblue4", na.value ="darkgrey", name="Percent (%)") +
  theme(axis.text =  element_blank(),
    axis.ticks =  element_blank(),
    panel.background = element_blank())
ggsave("blkAaPlot.png",blkAaPlot)
```

```{r}
asianPlot<-ggplot(data_sf)+geom_sf(aes(fill=PROP_ASIAN.percent), color = NA) + 
  labs(title = "Percent Asian Alone in US Census Tracts") + 
  scale_fill_gradient(low = "grey94", high = "chartreuse4", na.value ="darkgrey", name="Percent (%)") +  
  theme(axis.text =  element_blank(),
    axis.ticks =  element_blank(),
    panel.background = element_blank())
ggsave("asianPlot.png",asianPlot)
```

```{r}
nativeIslPlot<-ggplot(data_sf)+geom_sf(aes(fill=PROP_NATIVEISL.percent), color = NA) + 
  labs(title = "Percent Native Hawaiian and Other Pacific Islander Alone in US Census Tracts") + 
  scale_fill_gradient(low = "grey94", high = "red3", na.value ="darkgrey", name="Percent (%)") +  
  theme(axis.text =  element_blank(),
    axis.ticks =  element_blank(),
    panel.background = element_blank())
ggsave("nativeIslPlot.png",nativeIslPlot)
```

```{r}
nativePlot<-ggplot(data_sf)+geom_sf(aes(fill=PROP_NATIVE.percent), color = NA) + 
  labs(title = "Percent American Indian and Alaska Native Alone in US Census Tracts") + 
  scale_fill_gradient(low = "grey94", high = "darkolivegreen", na.value ="darkgrey", name="Percent (%)") +  
  theme(axis.text =  element_blank(),
    axis.ticks =  element_blank(),
    panel.background = element_blank())
ggsave("nativePlot.png",nativePlot)
```

```{r}
degPlot<-ggplot(data_sf)+geom_sf(aes(fill=PROP_BACHDEG_25.percent), color = NA) + 
  labs(title = "Percent 25+ with Bachelors in US Census Tracts") + 
  scale_fill_gradient(low = "grey94", high = "navy", na.value ="darkgrey", name="Percent (%)")+  theme(axis.text =  element_blank(),
    axis.ticks =  element_blank(),
    panel.background = element_blank())
ggsave("degPlot.png",degPlot)
```

```{r}
urbanPlot<-ggplot(data_sf)+geom_sf(aes(fill=PROP_URBAN.percent), color = NA) + 
  labs(title = "Percent Urban in US Census Tracts") + 
  scale_fill_gradient(low = "grey94", high = "tan2", na.value ="darkgrey", name="Percent (%)") +  
  theme(axis.text =  element_blank(),
    axis.ticks =  element_blank(),
    panel.background = element_blank())
ggsave("urbanPlot.png",urbanPlot)
```

```{r}
popDensPlot<-ggplot(data_sf)+geom_sf(aes(fill=POPDENS.ppl_km2), color = NA) + 
  labs(title = "Population Density (people/km2) US Census Tracts") + 
  scale_fill_gradient(low = "grey94", high = "red4", na.value ="darkgrey", name="Population Density (people/km2)") +  
  theme(axis.text =  element_blank(),
    axis.ticks =  element_blank(),
    panel.background = element_blank())
ggsave("popDensPlot.png",popDensPlot)
```

```{r}
logpopDensPlot<-ggplot(data_sf)+geom_sf(aes(fill=log_POPDENS.ln_ppl_km2), color = NA) + 
  labs(title = "Natural-Log Population Density US Census Tracts") + 
  scale_fill_gradient(low = "grey94", high = "red4", na.value ="darkgrey", name="ln(people/km2)") +  
  theme(axis.text =  element_blank(),
    axis.ticks =  element_blank(),
    panel.background = element_blank())
ggsave("logpopDensPlot.png",logpopDensPlot)
```

```{r}
adiPlot<-ggplot(data_sf)+geom_sf(aes(fill=TRACT_ADI.ave), color = NA) + 
  labs(title = "Average ADI in US Census Tracts") + 
  scale_fill_gradient(low = "grey94", high = "orangered3", na.value ="gray", name="ADI Score") +  
  theme(axis.text =  element_blank(),
    axis.ticks =  element_blank(),
    panel.background = element_blank())
ggsave("adiPlot.png",adiPlot)
```

```{r}
incomePlot<-ggplot(data_sf)+geom_sf(aes(fill=PERCAPINCOME_12mo.2010USD), color = NA) + 
  labs(title = "Average 12-Month Per Capita Income in US Census Tracts") + 
  scale_fill_gradient(low = "grey94", high = "green4", na.value ="gray", name="2010USD", labels = scales::comma) + theme(axis.text =  element_blank(),
    axis.ticks =  element_blank(),
    panel.background = element_blank())
ggsave("incomePlot.png",incomePlot)
```

# This is my code for the machine learning part. My code isn't working for the geometry part of the data, so only run either this code or Nora's code above if you don't want to mess up the code. If you want to see the outputs of machine learning, start from that section without running the intermediate code below. 

```{r}
plot_usmap("states",fill=NA) 
```

```{r}
tidycensus::census_api_key("cab9212edc0bd6f90c452f82768de2863bcaf428")
library(foreign)
shapefile <- read.dbf("US_tract_2010.dbf", as.is = FALSE)

UStract <- subset(shapefile, !(STATEFP10 %in% c("02", "15", "72")))
```

```{r}
# merge spatial, census, and ADI data
UStract$TRACTCE10 <- as.numeric(as.character(UStract$TRACTCE10))
names(UStract)[names(UStract) == "TRACTCE10"] <- "TRACTA"
UStract$TRACTA <- as.character(as.numeric(UStract$TRACTA))
names(UStract)[names(UStract) == "STATEFP10"] <- "STATEFP"
UStract$STATEFP <- as.character(as.numeric(UStract$STATEFP))
names(UStract)[names(UStract) == "GEOID10"] <- "GEOID"

UScopy <- data.frame(UStract)
UScopy$geometry <- paste(UStract$INTPTLAT10, UStract$INTPTLON10, sep = ",")

```

```{r}
data_sf <- merge(UScopy,data,by="GEOID")
```


# Start here. 

# Machine Learning- Sage: 

# Population Density looks weird, so I am altering the formula. Please see comments in that section. 

Decision Tree Code and Output: 

```{r warning= FALSE}
library(tidyflow)
# Define the decision tree and tell it the the dependent
# variable is continuous ('mode' = 'regression')
dt_ctrl <- rpart.control(maxdepth = 3)
mod1 <- set_engine(decision_tree(mode = "regression"), "rpart", control = dt_ctrl)
tflow <-
  # Plug the data
  data_sf %>%
  # Begin the tidyflow
  tidyflow(seed = 23151) %>%
  # Separate the data into training/testing (we are keeping 3/4 of the data for training)
  plug_split(initial_split, prop = 3/4) %>%
  # Plug the formula
  plug_formula(PM2.5.ug_m3_LC ~ TOTAL_POP.ppl + BACHDEG_MALE25.ppl + BACHDEG_FEM25.ppl + URBAN_POP.ppl + RURAL_POP.ppl + PERCAPINCOME_12mo.2010USD + CENSUS_REG + PROP_WH.percent + PROP_BLKAA.percent + PROP_NATIVE.percent + PROP_NATIVEISL.percent + PROP_ASIAN.percent + PROP_BACHDEG_25.percent + TRACT_ADI.ave + log(POPDENS.ppl_km2)) %>%
  # Plug the model
  plug_model(mod1)

res_rf <- tflow %>% fit()

vanilla_fit <- fit(tflow)
tree <- pull_tflow_fit(vanilla_fit)$fit
rpart.plot(tree)
```
Random Forest and Ranking of Variables: 

```{r}
data_sfnew <- na.omit(data_sf)
```

```{r}
# Define the random forest. There are about 15 predictor variables, I used 1/3 for mtry, so 5.
rf_mod <-
  rand_forest(mode = "regression", mtry = 5) %>%
  set_engine("ranger", importance = "impurity")

# Define the `tidyflow` with the random forest model
tflow3 <-
data_sfnew %>%
  tidyflow(seed = 1738) %>%
  plug_formula(PM2.5.ug_m3_LC ~ TOTAL_POP.ppl + BACHDEG_MALE25.ppl + BACHDEG_FEM25.ppl + URBAN_POP.ppl + RURAL_POP.ppl + PERCAPINCOME_12mo.2010USD + CENSUS_REG + PROP_WH.percent + PROP_BLKAA.percent + PROP_NATIVE.percent + PROP_NATIVEISL.percent + PROP_ASIAN.percent + PROP_BACHDEG_25.percent + TRACT_ADI.ave + log(POPDENS.ppl_km2)) %>%
  plug_split(initial_split, prop = 3/4) %>%
  plug_model(rf_mod)

res_rf3 <- tflow3 %>% fit()

res_rf3 %>%
  pull_tflow_fit() %>%
  .[['fit']] %>%
  vip()+
  theme_minimal()
```

Boosted Model and Ranking of Variable Importance: 

```{r}
boost_mod <-
  boost_tree(mode = "regression", trees = 500) %>%
  set_engine("xgboost")

tflow4 <-
  data_sfnew %>%
  tidyflow(seed = 51231) %>%
   plug_formula(PM2.5.ug_m3_LC ~ TOTAL_POP.ppl + BACHDEG_MALE25.ppl + BACHDEG_FEM25.ppl + URBAN_POP.ppl + RURAL_POP.ppl + PERCAPINCOME_12mo.2010USD + CENSUS_REG + PROP_WH.percent + PROP_BLKAA.percent + PROP_NATIVE.percent + PROP_NATIVEISL.percent + PROP_ASIAN.percent + PROP_BACHDEG_25.percent + TRACT_ADI.ave + log(POPDENS.ppl_km2)) %>%
  plug_split(initial_split, prop = 3/4) %>%
  plug_model(boost_mod)

res_boost <- tflow4 %>% fit()

res_boost %>%
  pull_tflow_fit() %>%
  .[['fit']] %>%
  vip()+
  theme_minimal()
```

```{r warning= FALSE, message= FALSE}

res_rf %>%
  predict_training() %>%
  rmse(PM2.5.ug_m3_LC, .pred)

res_rf %>%
  predict_testing() %>%
  rmse(PM2.5.ug_m3_LC, .pred)

res_rf3 %>%
  predict_training() %>%
  rmse(PM2.5.ug_m3_LC, .pred)

res_rf3 %>%
  predict_testing() %>%
  rmse(PM2.5.ug_m3_LC, .pred)

res_boost %>%
  predict_training() %>%
  rmse(PM2.5.ug_m3_LC, .pred)

res_boost %>%
  predict_testing() %>%
  rmse(PM2.5.ug_m3_LC, .pred)


#For Byron's part
#ctpred <- predict(carbon_modelmultiplelinearmodel, new_data = carbon_train) %>% 
  #as.data.frame() %>% 
  #pull(.pred)
#ctestpred <- predict(carbon_modelmultiplelinearmodel, new_data = carbon_test) %>% 
  #as.data.frame() %>% 
  #pull(.pred)

#lasttrain_rmse <- rmse_vec(carbon_train$res.carbon.pc.mt, ctpred)
#lasttest_rmse <- rmse_vec(carbon_test$res.carbon.pc.mt, ctestpred)

#lasttest_rmse
#lasttrain_rmse
```
Byron's Linear Model, converted to more easily take RMSE: 

```{r}
library(tidymodels)

set.seed(020304)

lr_spec <- linear_reg() %>%
  set_engine("lm") %>%
  set_mode("regression")

pm_modelmultiplelinearmodel <- lr_spec %>%
  fit(PM2.5.ug_m3_LC~PROP_BLKAA.percent+PROP_BACHDEG_25.percent+PERCAPINCOME_12mo.2010USD+PROP_URBAN.percent+ log(POPDENS.ppl_km2), data=data_sf)
(pm_modelmultiplelinearmodel)
```
```{r}
RMSE1 <- predict(pm_modelmultiplelinearmodel, new_data = data_sf) %>% 
  as.data.frame() %>% 
  pull(.pred)

linearmodelrmse <- rmse_vec(data_sf$PM2.5.ug_m3_LC, RMSE1)
linearmodelrmse
```

Code for Table: 

```{r}
# Code for Table: 

library(dplyr)
rmse_fortable <- data.frame(model = c("Decision Tree Training", "Decision Tree Test", "Random Forest Training", "Random Forest Test", "Boosted Training", "Boosted Test", "Linear Regression of Predictive Variables"),  rmse = c(2.117053, 2.129072, 0.6950791, 1.651716, 0.9260673, 1.728985, 2.655934))

rmse_sorted <- arrange(rmse_fortable, rmse)
print(rmse_sorted)
```

# New Part

```{r}
# linear model exploring unadjusted relationship between PM2.5 and proportion white
m1 <- lm(PM2.5.ug_m3_LC~ PROP_WH.percent, data=data_sf)
summary(m1)
```

```{r}
# linear model exploring association between PM2.5 and proportion white, controlling for education, per-capita income, and prop urban

m2 <- lm(PM2.5.ug_m3_LC~PROP_WH.percent+PROP_BACHDEG_25.percent+PERCAPINCOME_12mo.2010USD+PROP_URBAN.percent, data=data_sf)
summary(m2)
```
```{r}
mb <- lm(PM2.5.ug_m3_LC~PROP_BLKAA.percent+PROP_BACHDEG_25.percent+PERCAPINCOME_12mo.2010USD+PROP_URBAN.percent+ log(POPDENS.ppl_km2), data=data_sf)
summary(mb)
check_model(mb)
```

```{r}
check_model(m2)
```

```{r}
# reshape data to calculate population-weighted averages

data_long <- data %>% pivot_longer(cols=c(POPWH.ppl, POPBLKAA.ppl, POPNATIVE.ppl, POPASIAN.ppl, POPNATIVEISL.ppl), names_to = "racial_group", values_to = "count.ppl")
data_long$racial_group <- as.factor(data_long$racial_group)
```

```{r}
# calculate population-weighted averages

data_long <- data_long %>% group_by(racial_group) %>% mutate(sum(count.ppl))

data_long$frac_pm25 <- (data_long$count.ppl/data_long["sum(count.ppl)"])*data_long$PM2.5.ug_m3_LC

data_long <- as.data.frame(data_long) # correct issue with some sort of nesting of columns...

data_long %>% group_by(racial_group) %>% summarize(sum(frac_pm25))

```

```{r}
# fit linear model to check anova assumptions (with weights)

m3 <- lm(PM2.5.ug_m3_LC~racial_group, data = data_long, weights = count.ppl)
summary(m3)
```

```{r warning= FALSE}
ggqqplot(data_long, "PM2.5.ug_m3_LC", facet.by = "racial_group")
```

```{r}
# homogeneity of variance
plot(m3, 1)
```


```{r}
m4 <- aov(PM2.5.ug_m3_LC~racial_group, data = data_long, weights = count.ppl)
summary(m4)
```


```{r}
# logistic regression - a one-unit increase in ADI is associated with a small decrease in the odds of (beta = -8.09e-03, 95% CI [-9.02e-03, -7.17e-03], p < .001) a tract being in the top decile of PM2.5 concentrations
data <- data %>% mutate(ntile(PM2.5.ug_m3_LC, 10))

data$pm_90th <- ifelse(data["ntile(PM2.5.ug_m3_LC, 10)"] == 10, 1, 0)
#quantile(data$TRACT_ADI.ave) > 0.9

m5 <- glm(pm_90th ~ TRACT_ADI.ave, family = binomial, data = data)
summary(m5)
library(report)
report(m5)
```
# Byron's Model Converted to Linear regression




# Byron's linear regression and logistic regression analysis updated 4/21

## linear relationship between % race and PM2.5 concentrations

### exploration

```{r}
# linear model exploring unadjusted relationship between PM2.5 and proportion white
m1 <- lm(PM2.5.ug_m3_LC~PROP_WH.percent, data=data)
summary(m1)
```

```{r}
# linear model exploring association between PM2.5 and proportion white, controlling for education, per-capita income, and prop urban

m2 <- lm(PM2.5.ug_m3_LC~PROP_WH.percent+PROP_BACHDEG_25.percent+PERCAPINCOME_12mo.2010USD+PROP_URBAN.percent, data=data) # swap in log-population density for prop_urban
summary(m2)
```
```{r}
library(car)

# pm2.5 explained by % black
mb <- lm(PM2.5.ug_m3_LC~PROP_BLKAA.percent+PROP_BACHDEG_25.percent+PERCAPINCOME_12mo.2010USD+PROP_URBAN.percent+log_POPDENS.ln_ppl_km2, data=data)

# pm2.5 explained by % white
mw <- lm(PM2.5.ug_m3_LC~PROP_WH.percent+PROP_BACHDEG_25.percent+PERCAPINCOME_12mo.2010USD+PROP_URBAN.percent+log_POPDENS.ln_ppl_km2, data=data)

# pm2.5 explained by % asian
ma <- lm(PM2.5.ug_m3_LC~PROP_ASIAN.percent+PROP_BACHDEG_25.percent+PERCAPINCOME_12mo.2010USD+PROP_URBAN.percent+log_POPDENS.ln_ppl_km2, data=data)
```


### % black

```{r}
summary(mb)
mbt <- tidy(mb, conf.int = TRUE) %>% mutate_if(is.numeric, signif, digits=2)
mbt
avPlot(mb, variable = "PROP_BLKAA.percent", xlab = "% black | others", ylab = "PM2.5 (ug/m^3)", main="")
```

### % white

```{r}
summary(mw)
mwt <- tidy(mw, conf.int = TRUE) %>% mutate_if(is.numeric, signif, digits=2)
mwt
avPlot(mw, variable = "PROP_WH.percent", xlab = "% white | others", ylab = "PM2.5 (ug/m^3)", main="")
```

### % asian

```{r}
summary(ma)
mat <- tidy(ma, conf.int = TRUE) %>% mutate_if(is.numeric, signif, digits=2)
mat
avPlot(ma, variable = "PROP_ASIAN.percent", xlab = "% asian | others", ylab = "PM2.5 (ug/m^3)", main="")
```



```{r}
check_model(mb)
```

```{r}
# mb grouped by region

# Use the "group_by" function to group the data by region and fit a linear model for pm2.5 explained by race
mb_region <- data %>%
  nest_by(CENSUS_REG) %>%
  mutate(model = list(lm(PM2.5.ug_m3_LC~PROP_BLKAA.percent+PROP_BACHDEG_25.percent+PERCAPINCOME_12mo.2010USD+PROP_URBAN.percent+log_POPDENS.ln_ppl_km2, data = data))) %>%
  summarise(tidy(model, conf.int = TRUE))
mb_region <- mb_region %>% filter(term=="PROP_BLKAA.percent") %>% arrange(CENSUS_REG) %>% mutate_if(is.numeric, signif, digits=2) %>% relocate(conf.low, .after= estimate) %>% relocate(conf.high, .after = conf.low)
mb_region
#write_clip(mb_region)
```

## Population-weighted mean PM2.5 concenrtations

```{r}
# reshape data to calculate population-weighted averages

data_long <- data %>% pivot_longer(cols=c(POPWH.ppl, POPBLKAA.ppl, POPNATIVE.ppl, POPASIAN.ppl, POPNATIVEISL.ppl), names_to = "racial_group", values_to = "count.ppl")
data_long$racial_group <- as.factor(data_long$racial_group)
```

```{r}
# calculate population-weighted averages

data_long <- data_long %>% group_by(racial_group) %>% mutate(sum(count.ppl))

data_long$frac_pm25 <- (data_long$count.ppl/data_long["sum(count.ppl)"])*data_long$PM2.5.ug_m3_LC

data_long <- as.data.frame(data_long) # correct issue with some sort of nesting of columns...

data_long %>% group_by(racial_group) %>% summarize(sum(frac_pm25))

```

```{r}
# fit linear model to check anova assumptions (with weights)

m3 <- lm(PM2.5.ug_m3_LC~racial_group, data = data_long, weights = count.ppl)
summary(m3)
```

```{r}
# Compute Shapiro-Wilk test of normality (doesn't work because size of dataset)
shapiro_test(residuals(m3))
```

```{r}
ggqqplot(data_long, "PM2.5.ug_m3_LC", facet.by = "racial_group")
```

```{r}
# homogeneity of variance
plot(m3, 1)
```

## ANOVA (this didn't work; excluding for now)

```{r}
m4 <- aov(PM2.5.ug_m3_LC~racial_group, data = data_long, weights = count.ppl)
summary(m4)
```

```{r}
# Tukey post-hoc test doesn't work with population-weighted ANOVA

TukeyHSD(m4)
```

## Logistic regression: odds of high PM2.5 concentration, given increase in ADI score

```{r}
library(report)
# logistic regression - a one-unit increase in ADI is associated with a small decrease in the odds of (beta = -8.09e-03, 95% CI [-9.02e-03, -7.17e-03], p < .001) a tract being in the top decile of PM2.5 concentrations
data <- data %>% mutate(ntile(PM2.5.ug_m3_LC, 10))

data$pm_50th <- ifelse(data["ntile(PM2.5.ug_m3_LC, 10)"] > 5, 1, 0)
data$pm_90th <- ifelse(data["ntile(PM2.5.ug_m3_LC, 10)"] == 10, 1, 0)

m5 <- glm(pm_50th ~ TRACT_ADI.ave, family = binomial, data = data)
summary(m5)
report(m5)

m6 <- glm(pm_90th ~ TRACT_ADI.ave, family = binomial, data = data)
summary(m6)
report(m6)
```
```{r}
hist(data$pm_50th)
```

```{r}
plot <- ggplot(data, aes(x=TRACT_ADI.ave, y=pm_50th, color = CENSUS_REG)) + 
  geom_jitter(width = 0, height = 0.05) +
  geom_smooth(method="glm",  method.args = list(family="binomial")) + 
  labs(x = "ADI Score", y = "P(>50th %ile PM2.5)", title = "P(>50th %ile PM2.5) per ADI unit increase by region") +
  facet_wrap(~ CENSUS_REG) #+
  #scale_color_manual(values = c(fcolor, mcolor))
plot
ggsave("logistic regression.png")
```

```{r}
library(broom)
library(tidyverse)
# linear regression of pm2.5 explained by adi grouped by region 

# Use the "group_by" function to group the data by region and fit a linear model for pm2.5 explained by adi
model_region <- data %>%
  nest_by(CENSUS_REG) %>%
  mutate(model = list(lm(PM2.5.ug_m3_LC ~ TRACT_ADI.ave, data = data))) %>%
  summarise(tidy(model, conf.int = TRUE))
model_region <- model_region %>% arrange(CENSUS_REG) %>% filter(term=="TRACT_ADI.ave") %>% mutate_if(is.numeric, signif, digits=2) %>% relocate(conf.low, .after= estimate) %>% relocate(conf.high, .after = conf.low)
model_region
```

```{r}
library(clipr)
write_clip(model_region)
```



